---
output: github_document
editor_options: 
  chunk_output_type: console
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-"
)
```

[![CRAN Status](http://www.r-pkg.org/badges/version/ggmosaic)](https://cran.r-project.org/package=ggmosaic) [![CRAN RStudio mirror downloads](http://cranlogs.r-pkg.org/badges/ggmosaic)](http://www.r-pkg.org/pkg/ggmosaic) 
[![Travis-CI Build Status](https://travis-ci.org/haleyjeppson/ggmosaic.svg?branch=master)](https://travis-ci.org/haleyjeppson/ggmosaic)

# ggmosaic

ggmosaic was designed to create visualizations of categorical data and is capable of producing bar charts, stacked bar charts, mosaic plots, and double decker plots.

## Installation

You can install ggmosaic from github with:

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("haleyjeppson/ggmosaic")
```

## Example

```{r example}
library(ggmosaic)
ggplot(data = fly) +
  geom_mosaic(aes(x = product(rude_to_recline), fill=do_you_recline)) +
  theme_mosaic()
```

## geom_mosaic: setting the aesthetics
 
Aesthetics that can be set:

- `weight`: select a weighting variable

- `x`: select variables to add to formula  
    - declared as `x = product(x1, x2, ...)` 
    
- `fill`: select a variable to be filled  
    - if the variable is not also called in `x`, it will be added to the formula in the first position  
    
- `conds` : select a variable to condition on  
    - declared as  `conds = product(cond1, cond2, ...)`  

These values are then sent through `productplots` functions to create the formula for the desired distribution. The formula is constructed as:  `weight ~ fill + x | conds`


## Version compatibility issues with ggplot2

Since the initial release of ggmosaic, ggplot2 has evolved considerably. And as ggplot2 continues to evolve, ggmosaic must continue to evolve alongside it. Although these changes affect the underlying code and not the general usage of ggmosaic, the general user may need to be aware of compatibility issues that can arise between versions. The table below summarizes the compatibility between versions. 
  
Version 3.3.0 of ggplot2 included substantial changes to the implementation of the positional-guide internals ([see here](https://www.tidyverse.org/blog/2020/03/ggplot2-3-3-0/)). Because ggmosaic is implemented with a coordinate system unique to ggmosaic, `scale_x_productlist()` and `scale_y_productlist()`, the scales needed to be modified; the changes were implemented in ggmosaic 0.3.0.


| ggmosaic | ggplot2 | Axis labels                                                                   | Tick marks    |
|----------|---------|-------------------------------------------------------------------------------|---------------|
| 0.3.3    | 3.3.3   | x                                                                             | x             |
| 0.3.0    | 3.3.0   | x                                                                             | x             |
| 0.2.2    | 3.3.0   | Default labels are okay, but must use <br>`scale\_*\_productlist()` to modify | No tick marks |
| 0.2.2    | 3.2.0   | Default labels okay, but must use <br>`scale\_*\_productlist()` to modify     | x             |
| 0.2.0    | 3.2.0   | Default labels are wrong, but can use <br>`labs()` to modify                  | x             |

