---
output: github_document
editor_options: 
  chunk_output_type: console
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-"
)
```

[![CRAN Status](http://www.r-pkg.org/badges/version/ggmosaic)](https://cran.r-project.org/package=ggmosaic) [![CRAN RStudio mirror downloads](http://cranlogs.r-pkg.org/badges/ggmosaic)](http://www.r-pkg.org/pkg/ggmosaic) 
[![Travis-CI Build Status](https://travis-ci.org/haleyjeppson/ggmosaic.svg?branch=master)](https://travis-ci.org/haleyjeppson/ggmosaic)

# ggmosaic

ggmosaic was designed to create visualizations of categorical data and is capable of producing bar charts, stacked bar charts, mosaic plots, and double decker plots.

## Installation

You can install ggmosaic from github with:

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("haleyjeppson/ggmosaic")
```

## Example

```{r example}
library(ggmosaic)
ggplot(data = fly) +
  geom_mosaic(aes(x = product(rude_to_recline), fill=do_you_recline))
```

## geom_mosaic: setting the aesthetics
 
Aesthetics that can be set:

- `weight`: select a weighting variable

- `x`: select variables to add to formula  
    - declared as `x = product(x1, x2, ...)` 
    
- `fill`: select a variable to be filled  
    - if the variable is not also called in `x`, it will be added to the formula in the first position  
    
- `conds` : select a variable to condition on  
    - declared as  `conds = product(cond1, cond2, ...)`  

These values are then sent through `productplots` functions to create the formula for the desired distribution. The formula is constructed as:  `weight ~ fill + x | conds`
