---
title: "Mosaic plots with `ggplot2`"
author: "Haley Jeppson and Heike Hofmann"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
  toc: true
vignette: >
  %\VignetteIndexEntry{Mosaic plots with `ggplot2`}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---


  ```{r setup, echo=FALSE}

knitr::opts_chunk$set(fig.width = 6,
                      fig.height = 4,
                      fig.align='center',
                      dev = "png")

```

```{r echo=FALSE, message=FALSE}
library(ggmosaic)
# shouldn't be needed between these comments
library(plotly)
library(dplyr)
library(purrr)
library(tidyr)
# shouldn't be needed
library(ggplot2)
library(gridExtra)
library(grid)
data(NHANES, package="NHANES")
multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL) {
  library(grid)

  # Make a list from the ... arguments and plotlist
  plots <- c(list(...), plotlist)

  numPlots = length(plots)

  # If layout is NULL, then use 'cols' to determine layout
  if (is.null(layout)) {
    # Make the panel
    # ncol: Number of columns of plots
    # nrow: Number of rows needed, calculated from # of cols
    layout <- matrix(seq(1, cols * ceiling(numPlots/cols)),
                     ncol = cols, nrow = ceiling(numPlots/cols))
  }

  if (numPlots==1) {
    print(plots[[1]])

  } else {
    # Set up the page
    grid.newpage()
    pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout))))

    # Make each plot, in the correct location
    for (i in 1:numPlots) {
      # Get the i,j matrix positions of the regions that contain this subplot
      matchidx <- as.data.frame(which(layout == i, arr.ind = TRUE))

      print(plots[[i]], vp = viewport(layout.pos.row = matchidx$row,
                                      layout.pos.col = matchidx$col))
    }
  }
}
grid_arrange_shared_legend <- function(..., ncol = length(list(...)), nrow = 1, position = c("bottom", "right")) {
  
  plots <- list(...)
  position <- match.arg(position)
  g <- ggplotGrob(plots[[1]] + theme(legend.position = position))$grobs
  legend <- g[[which(sapply(g, function(x) x$name) == "guide-box")]]
  lheight <- sum(legend$height)
  lwidth <- sum(legend$width)
  gl <- lapply(plots, function(x) x + theme(legend.position="none"))
  gl <- c(gl, ncol = ncol, nrow = nrow)
  
  combined <- switch(position,
                     "bottom" = arrangeGrob(do.call(arrangeGrob, gl),
                                            legend,
                                            ncol = 1,
                                            heights = unit.c(unit(1, "npc") - lheight, lheight)),
                     "right" = arrangeGrob(do.call(arrangeGrob, gl),
                                           legend,
                                           ncol = 2,
                                           widths = unit.c(unit(1, "npc") - lwidth, lwidth)))
  grid.newpage()
  grid.draw(combined)
  
}
```

# Introduction

<!--
### Basic Explanation of `ggmosaic`  
  
`ggmosiac` was designed to create visualizations of categorical data, and has the capability to produce bar charts, stacked bar charts, mosaic plots, and double decker plots. The main focus of this vignette, however, will be on mosaic plots. A mosaic plot is a convenient graphical summary of the conditional distributions in a contingency table, and in a mosaic plot, the area of each graphical element is proportional to the underlying probability of that category. This allows us to easily visualize how the joint distribution is composed of the product of the conditional and marginal distributions -- which, in turn, allows us to see any association that may be occurring between the variables. Because the plot is constructed hierarchically, the ordering of the variables is very important.  There are many features that can be customized in `ggmosaic`. A few that will be discussed in this vignette are the types of partitioning available, the ordering, conditioning, and highlighting of the variables, and the spacing between the categories. -->

## Basic Explanation of `ggmosaic`  

- designed to create visualizations of categorical data
- can produce bar charts, stacked bar charts, mosaic plots, and double decker plots
- plots are constructed hierarchically, so the ordering of the variables is very important.
- integrated in ggplot2 as a `geom`
- allows for ggplot2 tools, such as facetting and layering


## Creation of `ggmosaic`

`ggmosaic` was created primarily using `ggproto` and the `productplots` package

`ggproto` allows you to extend ggplot2 from within your own packages 

- `ggmosaic` began as a geom extension of the `rect` geom
- used the data handling provided in the `productplots` package
- calculates xmin, xmax, ymin, and ymax for the `rect` geom to plot


## ggplot2 limitations

ggplot2 is not capable of handling a variable number of variables

- __current solution:__ read in the variables x1 and x2 as x = product(x1, x2)

    
- `product` function:
- creates a data frame that combines all of the variables listed 
- allows for it to pass `check_aesthetics` 
- then splits the variables back apart for the calculations
    

The product function creates limitiations for values the variables can take, and what the labels of variables can be. When the variables are combined, the values, variable name, and level are separated using ":", "-", and "."
    
- level-variable:value.level-variable:value
    
    
If any of the variable names or values of the variable contain one of those 3 symbols, the function will break
    
__Current solution:__ an option to redefine what symbols are used as separators
    
Default separators:
      
```{r echo=TRUE, tidy=TRUE, fig.align='center'}
    
get.separators() 
    
```
    
Separators redefined:
      
```{r eval=FALSE, echo=TRUE, tidy=TRUE, fig.align='center'}
  
set.separators(c(":", ";","|")) 
    
``` 
    
These limitations also lead to issues with the labeling, but those can be manually fixed.

## geom_mosaic: setting the aesthetics

Aesthetics that can be set:

- __weight__ : select a weighting variable
- __x__ : select variables to add to formula
    - declared as __x__ = product(x1, x2, ...)
- __fill__ : select a variable to be filled 
    - if the variable is not also called in __x__, it will be added to the formula in the first position
- __conds__ : select a variable to condition on

These values are then sent through `productplots` functions to create the formula for the desired distribution

  
 __Formula:__ **weight ~ fill + x | conds **
  

### From the aesthetics to the formula

Example of how the formula is built

- __weight__ = 1
- __x__ = product(Y, X)
- __fill__ = W
- __conds__ = Z
  
These aesthetics set up the formula for the distribution:

  
> __Formula:__ 1 ~ W + Y + X | Z 
  


Because a mosaic plot is constructed hierarchically through alternating spines, the ordering of the variables is very important.

## Weight ~ X

<!--The aesthetics set up the formula that determines the how the joint distribution will be broken down.-->


```{r formula-1b, message=FALSE, fig.align='center'}
set.separators(c(":", ";","|"))

 ggplot(data = NHANES) +
   geom_mosaic(aes(weight = Weight, x = product(SleepHrsNight), fill=factor(SleepHrsNight)), na.rm=TRUE) +
   labs(x="Hours of sleep a night ", title='f(SleepHrsNight)') + guides(fill=guide_legend(title = "SleepHrsNight", reverse = TRUE))
```


##  Weight ~ Y + X


```{r formula-2b, message=FALSE, fig.align='center'}
set.separators(c(":", ";","|"))

 ggplot(data = NHANES) +
   geom_mosaic(aes(weight = Weight, x = product(SleepHrsNight, AgeDecade), fill=factor(SleepHrsNight)), na.rm=TRUE) +    theme(axis.text.x=element_text(angle=-25, hjust= .1)) + labs(x="Age in Decades ", title='f(SleepHrsNight | AgeDecade) f(AgeDecade)') + guides(fill=guide_legend(title = "SleepHrsNight", reverse = TRUE))

```

##  Weight ~ X + Y / Z


```{r formula-4b, message=FALSE, fig.align='center'}
set.separators(c(":", ";","_"))

 ggplot(data = NHANES) +
   geom_mosaic(aes( x = product(SleepHrsNight, AgeDecade), fill=factor(SleepHrsNight), conds=product(Gender)), na.rm=TRUE, divider=mosaic("v")) +    theme(axis.text.x=element_text(angle=-25, hjust= .1)) + labs(x="Age in Decades ", title='f(SleepHrsNight, AgeDecade | Gender)') + guides(fill=guide_legend(title = "SleepHrsNight", reverse = TRUE))


```

##  Alternative to conditioning: facetting

<!--The aesthetics set up the formula that determines the how the joint distribution will be broken down.-->


```{r formula-5b, message=FALSE, fig.align='center'}
ggplot(data = NHANES) +
   geom_mosaic(aes( x = product(SleepHrsNight, AgeDecade), fill=factor(SleepHrsNight)), na.rm=TRUE) +    theme(axis.text.x=element_text(angle=-25, hjust= .1)) +
   labs(x="Age in Decades ", title='f(SleepHrsNight, AgeDecade | Gender)')  + facet_grid(Gender~.) + guides(fill=guide_legend(title = "SleepHrsNight", reverse = TRUE))


```


## Importance of ordering 

```{r order-b, message=FALSE, fig.align='center'}

order1 <- ggplot(data = NHANES) + geom_mosaic(aes(weight = Weight, x = product(SleepHrsNight, Gender), fill=factor(SleepHrsNight)), na.rm=TRUE, offset=0.015) + labs(x="Gender ", title='f(SleepHrsNight | Gender)  f(Gender)') + guides(fill=guide_legend(title = "SleepHrsNight", reverse = TRUE)) + theme(plot.title = element_text(size = rel(1)))

order2<- ggplot(data = NHANES) + geom_mosaic(aes(weight = Weight, x = product(Gender, SleepHrsNight), fill=factor(SleepHrsNight)), na.rm=TRUE, offset=0.015) + labs(x="", y="Gender", title='f(Gender | SleepHrsNight)  f(SleepHrsNight)') + guides(fill=guide_legend(title = "SleepHrsNight", reverse = TRUE)) + theme(plot.title = element_text(size = rel(1))) + coord_flip()

```

```{r order-3, message=FALSE, fig.width = 8, fig.height = 4, fig.align='center'}

grid_arrange_shared_legend(order1, order2, ncol = 2, nrow = 1, position = "right")


```



## Other features of geom_mosaic
  
Arguments unique to `geom_mosaic`:

- **divider:** used to declare the type of partitions to be used
- **offset:** sets the space between the first spine


## Divider function: Types of partitioning

Four options available for each partion:

- **vspine:**  <!-- (Vertical spine partition) --> width constant, height varies.
- **hspine:**  <!-- Horizontal spine partition: --> height constant, width varies.
- **vbar:**  <!--  Vertical bar partition: --> height constant, width varies.
- **hbar:**  <!--  Horizontal bar partition: --> width constant, height varies.
    


```{r partitions, message=FALSE, fig.width = 7, fig.height = 3.5}
set.separators(c(":", ";","|"))

a2 <- ggplot(data = NHANES) +
   geom_mosaic(aes( x = product(SleepHrsNight), fill=factor(SleepHrsNight)), divider="hbar", na.rm=TRUE) + theme(#axis.text.x=element_text(angle=35, hjust= 1),
     legend.position="none")+labs(x=" ", title='divider= "hbar"') + guides(fill=guide_legend(title = "SleepHrsNight", reverse = TRUE))


a1 <- ggplot(data = NHANES) +
   geom_mosaic(aes( x = product(SleepHrsNight), fill=factor(SleepHrsNight)), divider="hspine", na.rm=TRUE) + theme(#axis.text.x=element_text(angle=35, hjust= 1),
     legend.position="none") + labs(x=" ", title='divider= "hspine"') + guides(fill=guide_legend(title = "SleepHrsNight", reverse = TRUE))


b2 <- ggplot(data = NHANES) +
   geom_mosaic(aes( x = product(SleepHrsNight), fill=factor(SleepHrsNight)), divider="vbar", na.rm=TRUE) + theme(legend.position="none") + labs(y=" ", x="", title='divider= "vbar"') + guides(fill=guide_legend(title = "SleepHrsNight", reverse = TRUE))
 
b1 <- ggplot(data = NHANES) +
   geom_mosaic(aes(  x = product(SleepHrsNight), fill=factor(SleepHrsNight)), divider="vspine", na.rm=TRUE) + theme(legend.position="none") + labs(y=" ", x="", title='divider= "vspine"') + guides(fill=guide_legend(title = "SleepHrsNight", reverse = TRUE))

```


```{r plot, message=FALSE, fig.width = 8, fig.height = 4.5, fig.align='center'}

grid_arrange_shared_legend(a1, a2, b1, b2, ncol = 2, nrow = 2, position = "right")

```

## Partitioning with one or more variables

- __mosaic()__
    - default
    - will use spines in alternating directions 
    - begins with a horizontal spine 
- __mosaic("v")__
    - begins with a vertical spine and then alternates
- __ddecker()__
    - selects n-1 horizontal spines and ends with a vertical spine
- Define each type of partition 
    -  c("hspine", "vspine", "hbar")


```{r mosaic-a, message=FALSE, fig.width = 7, fig.height = 3.5}
set.separators(c(":", ";","|"))

m1 <-ggplot(data = NHANES) + geom_mosaic(aes(x=product(SleepHrsNight, Gender, AgeDecade), fill = factor(SleepHrsNight)), na.rm=T, divider=mosaic("h")) +
   theme(axis.text.x=element_blank(), legend.position="none")+labs(x=" ", title='divider= mosaic()') + guides(fill=guide_legend(title = "SleepHrsNight", reverse = TRUE))

m2 <-ggplot(data = NHANES) + geom_mosaic(aes(x=product(SleepHrsNight, Gender, AgeDecade), fill = factor(SleepHrsNight)), na.rm=T, divider=mosaic("v")) +
   theme(axis.text.x=element_blank(), legend.position="none")+labs(x=" ", title='divider= mosaic("v")') + guides(fill=guide_legend(title = "SleepHrsNight", reverse = TRUE))


m3 <-ggplot(data = NHANES) + geom_mosaic(aes(x=product(SleepHrsNight, Gender, AgeDecade), fill = factor(SleepHrsNight)), na.rm=T, divider=ddecker()) +
   theme(axis.text.x=element_blank(), legend.position="none")+labs(x=" ", title='divider= ddecker()') + guides(fill=guide_legend(title = "SleepHrsNight", reverse = TRUE))



```



```{r mosaic-3, message=FALSE, fig.width = 8, fig.height = 4.5, fig.align='center'}

grid_arrange_shared_legend(m1, m2, m3, ncol = 3, nrow = 1, position = "right")

```



```{r mosaic-d, message=FALSE, fig.width = 7, fig.height = 3.5}

m4 <-  ggplot(data = NHANES) + geom_mosaic(aes(x=product(SleepHrsNight, Gender, AgeDecade), fill = factor(SleepHrsNight)), na.rm=T, divider=c("vspine", "vspine", "hbar")) +
   theme(axis.text.x=element_blank(), legend.position="none")+labs(x=" ", title='divider= c("vspine", "vspine", "hbar")') + guides(fill=guide_legend(title = "SleepHrsNight", reverse = TRUE))

m5 <-  ggplot(data = NHANES) + geom_mosaic(aes(x=product(SleepHrsNight, Gender, AgeDecade), fill = factor(SleepHrsNight)), na.rm=T, divider=c("hbar", "vspine", "hbar")) +
   theme(axis.text.x=element_blank(), legend.position="none")+labs(x=" ", title='divider= c("hbar", "vspine", "hbar")') + guides(fill=guide_legend(title = "SleepHrsNight", reverse = TRUE))

m6 <-  ggplot(data = NHANES) + geom_mosaic(aes(x=product(SleepHrsNight, Gender, AgeDecade), fill = factor(SleepHrsNight)), na.rm=T, divider=c("hspine", "hspine", "hspine")) +
   theme(axis.text.x=element_blank(),
legend.position="none")+labs(x=" ", title='divider= c("hspine", "hspine", "hspine")') + guides(fill=guide_legend(title = "SleepHrsNight", reverse = TRUE))

m7 <-  ggplot(data = NHANES) + geom_mosaic(aes(x=product(SleepHrsNight, Gender, AgeDecade), fill = factor(SleepHrsNight)), na.rm=T, divider=c("vspine", "vspine", "vspine")) +
   theme(axis.text.x=element_blank(), legend.position="none")+labs(x=" ", title='divider= c("vspine", "vspine", "vspine")') + guides(fill=guide_legend(title = "SleepHrsNight", reverse = TRUE))


```


```{r mosaic-6, message=FALSE, fig.width = 8, fig.height = 4.5, fig.align='center'}

grid_arrange_shared_legend(m4, m5, m6, m7, ncol = 2, nrow = 2, position="right")

```



  
## geom_mosaic: offset

  
**offset:** Set the space between the first spine

- default = 0.01
- space between partitions decreases as layers build

## Adjusting the offset


```{r offset-b, message=FALSE}

oo1 <- ggplot(data = NHANES) +
  geom_mosaic(aes(weight = Weight,  x = product(Age), fill=factor(SleepHrsNight)), na.rm=TRUE) +   theme(axis.text.x=element_text(angle=0, hjust= .5))+labs(x="Age", y=" ",  title=" offset = 0.01") + guides(fill=guide_legend(title = "SleepHrsNight", reverse = TRUE)) 

o1 <- ggplot(data = happy) +
  geom_mosaic(aes(weight = wtssall,  x = product(age), fill=marital)) +
  theme(axis.text.x=element_text(angle=0, hjust= .5))+labs(x="Age", y=" ",  title=" offset = 0.01")+ guides(fill=guide_legend(title = "SleepHrsNight", reverse = TRUE)) 

oo2 <- ggplot(data = NHANES) +
  geom_mosaic(aes(weight = Weight,  x = product(Age), fill=factor(SleepHrsNight)), offset=0, na.rm=TRUE) +   theme(axis.text.x=element_text(angle=0, hjust= .5))+labs(x="Age", y=" ",  title=" offset = 0")+ guides(fill=guide_legend(title = "SleepHrsNight", reverse = TRUE)) 

o2 <- ggplot(data = happy) +
  geom_mosaic(aes(weight = wtssall,  x = product(age), fill=marital), offset = 0) +
  theme(axis.text.x=element_text(angle=0, hjust= .5))+labs(x="Age", y=" ",  title=" offset = 0") + guides(fill=guide_legend(title = "SleepHrsNight", reverse = TRUE))


```

```{r offset-plot, message=FALSE, fig.width = 8, fig.height = 4, fig.align='center'}

grid_arrange_shared_legend(oo1, oo2, nrow = 1, ncol =2, position="right")

```


## Plotly


```{r plotly, message=FALSE, echo=TRUE, tidy=TRUE, eval=FALSE}


gg <-ggplot(data = NHANES) + geom_mosaic(aes(x=product(SleepHrsNight, Gender, AgeDecade), fill = factor(SleepHrsNight)), na.rm=T, divider=ddecker(), offset = 0.025) 
```

```{r plotly-2, message=FALSE, echo=FALSE, tidy=TRUE}


gg <-ggplot(data = NHANES) + geom_mosaic(aes(x=product(SleepHrsNight, Gender, AgeDecade), fill = factor(SleepHrsNight)), offset=0.025, na.rm=T, divider=ddecker()) + guides(fill=guide_legend(title = "SleepHrsNight", reverse = TRUE))

```

```{r plotly-3, message=FALSE, echo=TRUE, tidy=TRUE}
# just for now commented out
# ggplotly(gg)


```




