---
title: "Mosaic plots with `ggplot2`"
author: "Haley Jeppson and Heike Hofmann"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Mosaic plots with `ggplot2`}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup, echo=FALSE}

knitr::opts_chunk$set(fig.width = 6,
                      fig.height = 4,
                      fig.align='center',
                      dev = "png")

```

```{r pkgs, echo=FALSE, message=FALSE}
library(ggmosaic)
library(gridExtra)
library(grid)
```

```{r data, echo=FALSE, message=FALSE}
data(fly)
my_pal <- c("#D43F3AFF","#EEA236FF", "#5CB85CFF", "#46B8DAFF", "#357EBDFF", "#9632B8FF", "#B8B8B8FF") #source: ggsci::pal_locuszoom("default")(7)

scale_fill_discrete <- function(...) scale_fill_manual(..., values = my_pal)

# create theme
theme_mosaic <- function (base_size = 12, base_family = "Helvetica") { 
  theme_gray() +
    theme(
      rect = element_rect(fill = NA, linetype = 0, colour = NA, size = base_size/22), 
      text = element_text(family = base_family, face = "plain", color = "black", size = base_size, hjust = 0.5, vjust = 0.5, angle = 0, lineheight = 0.9, margin = margin(), debug = FALSE),
      panel.grid.major.y = element_line(colour = "#D8D8D8"), 
      panel.grid.minor.y = element_blank(), 
      panel.grid.major.x = element_blank(), 
      panel.grid.minor.x = element_blank(), 
      panel.border = element_blank(), 
      panel.background = element_blank(), 
      legend.position = "right", 
      legend.key = element_rect(fill = "#FFFFFF00"),
      
      axis.title = element_text(size = rel(.8)), 
      axis.text = element_text(size = rel(.6)),
      legend.text = element_text(size = rel(.7)),
      legend.title = element_text(size = rel(.8)),
      plot.title = element_text(size = rel(.8)),
      plot.subtitle = element_text(size = rel(.8)),
      legend.key.size =    unit(.8, "lines")
    )
}

# set theme
theme_set(theme_mosaic())

grid_arrange_shared_legend <- function(..., ncol = length(list(...)), nrow = 1, position = c("bottom", "right")) {
  
  plots <- list(...)
  position <- match.arg(position)
  g <- ggplotGrob(plots[[1]] + theme(legend.position = position))$grobs
  legend <- g[[which(sapply(g, function(x) x$name) == "guide-box")]]
  lheight <- sum(legend$height)
  lwidth <- sum(legend$width)
  gl <- lapply(plots, function(x) x + theme(legend.position="none"))
  gl <- c(gl, ncol = ncol, nrow = nrow)
  
  combined <- switch(position,
                     "bottom" = arrangeGrob(do.call(arrangeGrob, gl),
                                            legend,
                                            ncol = 1,
                                            heights = unit.c(unit(1, "npc") - lheight, lheight)),
                     "right" = arrangeGrob(do.call(arrangeGrob, gl),
                                           legend,
                                           ncol = 2,
                                           widths = unit.c(unit(1, "npc") - lwidth, lwidth)))
  grid.newpage()
  grid.draw(combined)
  
}
```

# Introduction to `ggmosaic`  



- designed to create visualizations of categorical data
- can produce bar charts, stacked bar charts, mosaic plots, and double decker plots
- plots are constructed hierarchically, so the ordering of the variables is very important.
- integrated in ggplot2 as a `geom` which allows for facetting and layering

`ggmosaic` was created primarily using `ggproto` and the `productplots` package. `ggproto` allows you to extend ggplot2 from within your own packages.

- `ggmosaic` began as a geom extension of the `rect` geom
- used the data handling provided in the `productplots` package
- calculates xmin, xmax, ymin, and ymax for the `rect` geom to plot


## ggplot2 limitations

ggplot2 is not capable of handling a variable number of variables

- __current solution:__ read in the variables x1 and x2 as x = product(x1, x2)

- `product` function:
- a wrapper function for a list
- allows for it to pass `check_aesthetics` 

## geom_mosaic: setting the aesthetics

Aesthetics that can be set:

- __weight__ : select a weighting variable
- __x__ : select variables to add to formula
    - declared as __x__ = product(x1, x2, ...)
- __fill__ : select a variable to be filled 
    - if the variable is not also called in __x__, it will be added to the formula in the first position
- __conds__ : select a variable to condition on
    - declared as __conds__ = product(cond1, cond2, ...)

These values are then sent through `productplots` functions to create the formula for the desired distribution

  
 __Formula:__ **weight ~ fill + x | conds **
  

### From the aesthetics to the formula

Example of how the formula is built

- __weight__ = 1
- __x__ = product(Y, X)
- __fill__ = W
- __conds__ = product(Z)
  
These aesthetics set up the formula for the distribution:

  
> __Formula:__ 1 ~ W + X + Y | Z 
  


Because a mosaic plot is constructed hierarchically through alternating spines, the ordering of the variables is very important.

## 1 ~ X


```{r formula-1b, message=FALSE, fig.align='center'}
 ggplot(data = fly) +
  geom_mosaic(aes(x = product(rude_to_recline), fill=rude_to_recline), na.rm=TRUE) + 
  labs(title='f(rude_to_recline)')
```


##  1 ~ Y + X


```{r formula-2b, message=FALSE, fig.align='center'}

ggplot(data = fly) +
  geom_mosaic(aes(x = product(do_you_recline, rude_to_recline), fill=do_you_recline), na.rm=TRUE) + 
  labs(title='f(do_you_recline | rude_to_recline) f(rude_to_recline)')

```

##  1 ~ X + Y / Z


```{r formula-4b, message=FALSE, fig.align='center'}
ggplot(data = fly) +
  geom_mosaic(aes(x = product(do_you_recline, rude_to_recline), fill=do_you_recline, conds=product(gender)), na.rm=TRUE, divider=mosaic("v")) +
  labs(title='f(do_you_recline, rude_to_recline| gender)')
```

##  Alternative to conditioning: facetting



```{r formula-5b, message=FALSE, fig.align='center'}
ggplot(data = fly) +
  geom_mosaic(aes(x = product(do_you_recline, rude_to_recline), fill=do_you_recline), na.rm=TRUE) +
  labs(title='f(do_you_recline, rude_to_recline| gender)') + 
  facet_grid(gender~.)
```


## Importance of ordering 

```{r order-b, message=FALSE, fig.width = 7, fig.height = 2.4, fig.align='center'}
order1 <- ggplot(data = fly) +
  geom_mosaic(aes(x = product(do_you_recline, rude_to_recline), fill=do_you_recline), na.rm=TRUE)

order2 <- ggplot(data = fly) +
  geom_mosaic(aes(x = product(rude_to_recline, do_you_recline), fill=do_you_recline), na.rm=TRUE) + 
  coord_flip()

grid_arrange_shared_legend(order1, order2, ncol = 2, nrow = 1, position = "right")
```


## Other features of geom_mosaic
  
Arguments unique to `geom_mosaic`:

- **divider:** used to declare the type of partitions to be used
- **offset:** sets the space between the first spine


## Divider function: Types of partitioning

Four options available for each partion:

- **vspine:**  <!-- (Vertical spine partition) --> width constant, height varies.
- **hspine:**  <!-- Horizontal spine partition: --> height constant, width varies.
- **vbar:**  <!--  Vertical bar partition: --> height constant, width varies.
- **hbar:**  <!--  Horizontal bar partition: --> width constant, height varies.
    


```{r partitions, message=FALSE, fig.width = 7, fig.height = 4.8, fig.align='center'}
hbar <- ggplot(data = fly) +
  geom_mosaic(aes(x = product(rude_to_recline), fill=rude_to_recline), divider="hbar", na.rm=TRUE) + 
  labs(title='divider = "hbar"') 

hspine <- ggplot(data = fly) +
  geom_mosaic(aes(x = product(rude_to_recline), fill=rude_to_recline),  divider="hspine", na.rm=TRUE) + 
  labs(x=" ", title='divider = "hspine"') 

vbar <- ggplot(data = fly) +
  geom_mosaic(aes(x = product(rude_to_recline), fill=rude_to_recline), divider="vbar", na.rm=TRUE) + 
  labs(y=" ", x="", title='divider = "vbar"') 
 
vspine <- ggplot(data = fly) +
  geom_mosaic(aes(x = product(rude_to_recline), fill=rude_to_recline), divider="vspine", na.rm=TRUE) + 
  labs(y=" ", x="", title='divider = "vspine"') 

grid_arrange_shared_legend(hbar, hspine, vbar, vspine, ncol = 2, nrow = 2, position = "bottom")

```

## Partitioning with one or more variables

- __mosaic()__
    - default
    - will use spines in alternating directions 
    - begins with a horizontal spine 
- __mosaic("v")__
    - begins with a vertical spine and then alternates
- __ddecker()__
    - selects n-1 horizontal spines and ends with a vertical spine
- Define each type of partition 
    -  c("hspine", "vspine", "hbar")


```{r mosaic-a, message=FALSE, fig.width = 7, fig.height = 4.8, fig.align='center'}
h_mosaic <- ggplot(data = fly) +
  geom_mosaic(aes(x = product(do_you_recline, rude_to_recline, gender), fill=do_you_recline), na.rm=T, divider=mosaic("h")) +
  theme(axis.text.x=element_blank(), legend.position="none") + 
  labs(x=" ", title='divider= mosaic()')

v_mosaic <- ggplot(data = fly) +
  geom_mosaic(aes(x = product(do_you_recline, rude_to_recline, gender), fill=do_you_recline), na.rm=T, divider=mosaic("v")) +
  theme(axis.text.x=element_blank()) + 
  labs(x=" ", title='divider= mosaic("v")')

doubledecker <- ggplot(data = fly) +
  geom_mosaic(aes(x = product(do_you_recline, rude_to_recline, gender), fill=do_you_recline), na.rm=T, divider=ddecker()) +
  theme(axis.text.x=element_blank()) +
  labs(x=" ", title='divider= ddecker()')

grid_arrange_shared_legend(h_mosaic, v_mosaic, doubledecker, ncol = 2, nrow = 2, position = "bottom")
```



```{r mosaic-d, message=FALSE, fig.width = 7, fig.height = 5.5, fig.align='center'}
mosaic4 <-  ggplot(data = fly) +
  geom_mosaic(aes(x = product(do_you_recline, rude_to_recline, gender), fill=do_you_recline), na.rm=T, divider=c("vspine", "vspine", "hbar")) +
  theme(axis.text.y=element_blank()) +
  labs(x=" ", title='divider= c("vspine", "vspine", "hbar")') 

mosaic5 <-  ggplot(data = fly) +
  geom_mosaic(aes(x = product(do_you_recline, rude_to_recline, gender), fill=do_you_recline), na.rm=T, divider=c("hbar", "vspine", "hbar")) +
  theme(axis.text.x=element_blank()) +
  labs(x=" ", title='divider= c("hbar", "vspine", "hbar")')

mosaic6 <-  ggplot(data = fly) +
  geom_mosaic(aes(x = product(do_you_recline, rude_to_recline, gender), fill=do_you_recline), na.rm=T, divider=c("hspine", "hspine", "hspine")) +
  theme(axis.text.x=element_blank()) +
  labs(x=" ", title='divider= c("hspine", "hspine", "hspine")')

mosaic7 <-  ggplot(data = fly) +
  geom_mosaic(aes(x = product(do_you_recline, rude_to_recline, gender), fill=do_you_recline), na.rm=T, divider=c("vspine", "vspine", "vspine")) +
  theme(axis.text.y=element_blank()) +
  labs(x=" ", title='divider= c("vspine", "vspine", "vspine")')

grid_arrange_shared_legend(mosaic4, mosaic5, mosaic6, mosaic7, ncol = 2, nrow = 2, position="bottom")
```

  
## geom_mosaic: offset

  
**offset:** Set the space between the first spine

- default = 0.01
- space between partitions decreases as layers build

## Adjusting the offset


```{r offset-b, message=FALSE, fig.width = 7, fig.height = 4, fig.align='center'}

offset1 <- ggplot(data = fly) +
  geom_mosaic(aes(x = product(flight_freq, region), fill=flight_freq), na.rm=TRUE) + 
  labs(x="region", y=" ",  title=" offset = 0.01") +
  scale_y_productlist(name = "") +
  theme(axis.text.y=element_blank(),
        axis.text.x = element_text(angle = 90))

offset0 <- ggplot(data = fly) +
  geom_mosaic(aes(x = product(flight_freq, region), fill=flight_freq), na.rm=TRUE, offset = 0) + 
  labs(x="region", y=" ",  title=" offset = 0") +
  scale_y_productlist(name = "") +
  theme(axis.text.y=element_blank(),
        axis.text.x = element_text(angle = 90))

offset2 <- ggplot(data = fly) +
  geom_mosaic(aes(x = product(flight_freq, region), fill=flight_freq), na.rm=TRUE, offset = 0.02) + 
  labs(x="region", y=" ",  title=" offset = 0.02") +
  scale_y_productlist(name = "") +
  theme(axis.text.y = element_blank(),
        axis.text.x = element_text(angle = 90))

grid_arrange_shared_legend(offset0, offset1, offset2, nrow = 1, ncol =3, position="bottom")
```


## Current Issues

- `ggplotly()` no longer works with `geom_mosaic`


- There is an issue with labeling the axes. For example, to relabel the x-axis you will need to use `scale_x_productlist(name = "new label")` instead of `labs(x = "new label")`. See example below.

```{r, message=FALSE, fig.width = 7, fig.height = 2.4, fig.align='center'}
example1 <- ggplot(data = fly) +
  geom_mosaic(aes(x = product(rude_to_recline), fill=do_you_recline), na.rm=TRUE) + 
  labs(x = "Is it rude to recline?", title = "Relabeling does not work")

example2 <- ggplot(data = fly) +
  geom_mosaic(aes(x = product(rude_to_recline), fill=do_you_recline), na.rm=TRUE) + 
  labs(title = "Relabeling works") +
  scale_x_productlist(name = "Is it rude to recline?") +
  scale_y_productlist(name = "Do you recline?")

grid_arrange_shared_legend(example1, example2, nrow = 1, ncol = 2, position="right")
```


